---
title: "Project 3"
author: "Nancy Truong"
date: "2025-05-16"
output: html_document
---

# Alt.2 Ordinal logistic regression
Divide the plasma β-carotene concentrations into more than two categories with increasing concentration levels, using the function cut().

Divide the plasma β-carotene concentrations into more than two categories with increasing concen-Lec.11
tration levels, using the function cut().

```{r}
library(tidyverse); theme_set(theme_bw() + theme(text = element_text(size = 18)))
getwd()  # Check current working directory
list.files("Data")  # Check if 'carotene.xlsx' is inside the "Data" folder
#install.packages("readxl")  # Only run once if not installed
library(readxl)
library(dplyr)  # Load dplyr for select()
# Lecture 3 - Elasticity - multiple linear regression
library(caret)
library(pROC)
library(car)
library(rstatix)
library(GGally)
library(MASS)
data <- read_excel("Data/carotene.xlsx")
head(df)
library(nnet)
```
```{r}
#cut(df$betaplasma, breaks = c(0, 50, 200, 1600), include.lowest = TRUE)
```
Model the probabilities for the categories using ordinal logistic regression. Use the usual model selection tools to find a good model. You can investigate the Goodness-of-fit comparing the predictions to the true outcome

You could also ignore the fact that the categories are ordered and model the probabilities using a
multinomial logistic regression instead and compare the results with the ones from the ordinal logistic
model
```{r}
data |> mutate(
  plasma_cut = cut(betaplasma, breaks = c(0, 50, 225, 1600), include.lowest = TRUE),
  plasma_123 = factor(
    plasma_cut,
    levels = c("[0,50]", "(50,225]", "(225,1.6e+03]"),
    labels = c(1,2,3)
  )
) -> data
glimpse(data)
#view(data)
summary(data)
```

```{r}
mutate(data,
       sex = factor(sex,
                         levels = c(1, 2),
                         labels = c("male", "female")),
       smokstat = factor(smokstat,
                     levels = c(1, 2, 3),
                     labels = c("never", "former", "current")),
       vituse = factor(vituse,
                         levels = c(1, 2, 3),
                         labels = c("often", "notoften", "no"))
) -> data

data <- mutate(data,
               smokstat = relevel(smokstat, "never"), 
               sex = relevel(sex, "female"))
table(data$vituse)

```


```{r}
model.null <- polr(plasma_123 ~ 1, data = data)
sum.null <- summary(model.null)
sum.null
```

```{r}
model.full <- polr(plasma_123 ~ bmi + age + calories + fat + cholesterol + fiber + alcohol + betadiet + smokstat + sex + vituse, data = data)
sum.full <- summary(model.full)
sum.full
```
```{r}
vif(model.full)
sum.full
```

calories will be excluded
```{r}
model.excl <- polr(plasma_123 ~ bmi + age + fat + cholesterol + fiber + alcohol + betadiet + smokstat + sex + vituse, data = data)
sum.excl <- summary(model.excl)
sum.excl
```
Perform one Backwards elimination, starting with the full model, and one Forward selection, starting with the null model. Both versions should have the null model as lower scope and the full model as upper scope, and use BIC as selection criterion.

```{r}
cbind(beta = model.excl$coefficients, 
      expbeta = exp(model.excl$coefficients),
      exp(confint(model.excl)))
```
```{r}
cbind(zeta = model.excl$zeta, 
      expzeta = exp(model.excl$zeta))
```
```{r}
sum.excl$coefficients
```
Odds and probability for deficient, all x=0
```{r}
odds <- exp(model.excl$zeta["1|2"])
cbind(odds, p = odds/(1 + odds))
```
```{r}
odds <- exp(model.excl$zeta["2|3"])
cbind(odds, p = odds/(1 + odds))
```

Odds and probability for Fail, vituse = no

```{r}
odds <- exp(model.excl$zeta["1|2"])/exp(model.excl$coefficients["vituseno"])
cbind(odds, p = odds/(1 + odds))
```
```{r}
model.back <- step(model.excl, 
     scope = list(lower = model.null, upper = model.excl),
     direction = "backward")
sum.back <- summary(model.back)
```

Then, use each of the resulting models as starting model in a Stepwise regression, again with null and full models as scope, and BIC as criterion

```{r}
model.backstep <- step(model.back, 
     scope = list(lower = model.null, upper = model.excl),
     direction = "both")

```

```{r}
model.for <- step(model.null,
     
     scope = list(lower = model.null, upper = model.excl), 
     direction = "forward")
```
```{r}
model.forstep <- step(model.for, 
     scope = list(lower = model.null, upper = model.excl),
     direction = "both")

```
```{r}
mutate(data, 
       vituse_new = factor(vituse == "no",
                         levels = c(FALSE, TRUE),
                         labels = c("yes", "no"))) -> data
glimpse(data)
count(data, vituse_new)
# not vituseoften = often + notoften as reference
data <- mutate(data, vituse_new = relevel(vituse_new, "yes"))
#model.1 <- model.back
#sum.1 <- sum.backstep
model.back.new <- polr(formula = plasma_123 ~ vituse_new + betadiet + bmi + cholesterol + smokstat + 
    age, data = data)
sum.back.new <- summary(model.back.new)
sum.back.new
sum.back
```
LR-test

```{r}
D_diff <- sum.back.new$deviance - sum.back$deviance
df_diff <- sum.back.new$df.residual - sum.back$df.residual
cbind(D_diff, df_diff)

chi2_alpha <- qchisq(1 - 0.05, df_diff)
Pvalue <- pchisq(D_diff, df_diff, lower.tail = FALSE)
cbind(D_diff, df_diff, chi2_alpha, Pvalue)
#forward_stepwise_new_sum$deviance


```
large p-value, delete vitusenotoften

form now on, use vituse_new

```{r}
mutate(data, 
       smokstat_new = factor(smokstat == "never",
                         levels = c(FALSE, TRUE),
                         labels = c("yes", "no"))) -> data
#view(data)
count(data, smokstat_new)
# not vituseoften = often + notoften as reference
data <- mutate(data, smokstat_new = relevel(smokstat_new, "yes"))
#model.1 <- model.back
#sum.1 <- sum.backstep
model.back2.new <- polr(formula = plasma_123 ~ vituse_new + betadiet + bmi + cholesterol + smokstat_new + 
    age, data = data)
sum.back2.new <- summary(model.back2.new)
sum.back2.new
#sum.back
```


not sure what to comare AIC, BIC?
```{r}
anova(model.back.new, model.back2.new)
```
move on with 2 cat for smokstat
model.back2.new

```{r}
data |> mutate(yhat = predict(model.for)) -> pred.for
pred.for

confusionMatrix(data = pred.for$yhat,
                reference = pred.for$plasma_123)
```


```{r}
#excl_lm <- lm(betaplasma ~ (bmi + age + fat + cholesterol + fiber + alcohol + betadiet + smokstat + sex + vituse)^2, data = df)
#summary(excl_lm)
model.p2aic <- polr(formula = plasma_123 ~ bmi + age + fat + cholesterol + 
    fiber + betadiet + fat:betadiet + cholesterol:betadiet + 
    fiber:betadiet + bmi:betadiet, data = data)
sum.p2aic <- summary(model.p2aic)
```
```{r}
model.p2aicstep <- step(model.p2aic,
     
     scope = list(lower = model.null), 
     direction = "both")
sum.p2aicstep <- summary(model.p2aicstep)
sum.p2aicstep
```

```{r}
model.p2bic <- polr(formula = plasma_123 ~ betadiet + bmi + vituse_new + age + 
    betadiet:bmi, data = data)
model.p2bicstep <- step(model.p2bic,
     
     scope = list(lower = model.null), 
     direction = "both")
```



```{r}
data |> mutate(yhat = predict(forstep_model_step)) -> pred.forstep
#pred.forstep

confusionMatrix(data = pred.forstep$yhat,
                reference = pred.forstep$plasma_123)
```


```{r}
D_0 <- model.null$deviance
df_0 <- model.null$edf

info <- cbind(
  aic = AIC(model.null, model.excl, model.back, model.back2.new, model.p2aic, model.p2bic),
  bic = BIC(model.null, model.excl, model.back, model.back2.new, model.p2aic, model.p2bic),
  R2D = c(
    1 - 1, 
    1 - model.excl$deviance / D_0,
    1 - model.back$deviance / D_0,
    1 - model.back2.new$deviance / D_0,
    1 - model.p2aic$deviance / D_0,
    1 - model.p2bic$deviance
  ),
  R2D.adj = c(
    1 - 1, 
    1 - (model.excl$deviance + model.excl$edf - df_0) / D_0,
    1 - (model.back$deviance + model.back$edf - df_0) / D_0,
    1 - (model.back2.new$deviance + model.back2.new$edf - df_0) / D_0,
    1 - (model.p2aic$deviance + model.p2aic$edf - df_0) / D_0,
    1 - (model.p2bic$deviance + model.p2bic$edf - df_0) / D_0
  )
)

round(info, digits = 3)
```
aic: model.back2.new
bic: model.back2.new	
```{r}
# Predict classes
pred.null <- predict(model.null, data)
pred.excl <- predict(model.excl, data)
pred.back <- predict(model.back, data)
pred.back2.new <- predict(model.back2.new, data)
pred.p2aic <- predict(model.p2aic, data)
pred.p2bic <- predict(model.p2bic, data)

cm.null <- confusionMatrix(pred.null, data$plasma_123)
cm.excl <- confusionMatrix(pred.excl, data$plasma_123)
cm.back <- confusionMatrix(pred.back, data$plasma_123)
cm.back2.new <- confusionMatrix(pred.back2.new, data$plasma_123)
cm.p2aic <- confusionMatrix(pred.p2bic, data$plasma_123)
cm.p2bic <- confusionMatrix(pred.p2bic, data$plasma_123)
```


```{r}
cm.null
```
```{r}
cm.excl
```


```{r}
cm.back

```
```{r}
cm.back2.new

```

```{r}
cm.p2aic

```

```{r}
cm.p2bic
```
Acc: p2bic = p2aic = back =0.7238  
No Information Rate : 0.673
Kappa : back = 0.2655   

```{r}
null_model <- multinom(plasma_123 ~ 1, data = data)
null_sum <- summary(null_model)
null_sum
```
```{r}
table(data$plasma_123)
data |> mutate(plasma_123 = relevel(plasma_123, ref = 2)) ->
 data
```

```{r}
full_model <- multinom(plasma_123 ~ bmi + age + calories + fat + cholesterol + fiber + alcohol + betadiet + smokstat + sex + vituse, data = data)
vif(full_model)
```
```{r}
excl_model <- multinom(plasma_123 ~ bmi + age  + fat + cholesterol + fiber + alcohol + betadiet + smokstat + sex + vituse, data = data)
```

```{r}
aic_model <- step(null_model,
                    scope = list(upper = excl_model,
                                 lower = null_model),
                    direction = "both")
aic_sum <- summary(aic_model)
aic_sum
```
```{r}
bic_model <- step(null_model,
                     scope = list(upper = excl_model,
                                 lower = null_model),
                     direction = "both",
                     k = log(nrow(data)))
bic_sum <- summary(bic_model)
```

```{r}
aic_sum
```
```{r}
aic_new_model <- multinom(formula = plasma_123 ~ vituse_new + betadiet + bmi + age + 
    cholesterol + smokstat, data = data)
aic_new_sum <- summary(aic_new_model)

anova(aic_new_model, aic_model)
```

```{r}
 aic_new2_model <- multinom(formula = plasma_123 ~ vituse_new + betadiet + bmi + age + 
    cholesterol + smokstat_new, data = data)
aic_new2_sum <- summary(aic_new2_model)

anova(aic_new2_model, aic_new_model)
```
```{r}
bic_sum
```
```{r}
bic_new_model <- multinom(formula = plasma_123 ~ vituse_new + betadiet + bmi + age + 
    cholesterol, data = data)
bic_new_sum <- summary(bic_new_model)

anova(bic_new_model, bic_model)
```


```{r}
p2aic_model <- multinom(plasma_123 ~ bmi + age + fat + cholesterol + 
    fiber + betadiet + fat:betadiet + cholesterol:betadiet + 
    fiber:betadiet + bmi:betadiet, data = data)


```
```{r}
step(p2aic_model, 
     scope = list(lower = null_model),
     direction = "both")
```


```{r}
p2bic_model <- multinom(plasma_123 ~ betadiet + bmi + vituse_new + age + 
    betadiet:bmi, data = data)

step(p2bic_model, 
     scope = list(lower = null_model),
     direction = "both")
```

```{r}
D_0 <- model.null$deviance
df_0 <- model.null$edf
info2 <- cbind(aic = AIC(excl_model, aic_new2_model,bic_new_model , p2aic_model, p2bic_model),
              bic = BIC(excl_model, aic_new2_model,bic_new_model , p2aic_model, p2bic_model))
info2$r2 <- round(c(
  1 - excl_model$deviance/D_0,
  1 - aic_new2_model$deviance/D_0,
  1 - bic_new_model$deviance/D_0,
  1 - p2aic_model$deviance/D_0,
  1 - p2bic_model$deviance/D_0), digits = 3)

info2$r2.adj <- round(c(
  1 - (excl_model$deviance + (excl_model$edf - df_0))/D_0,
  1 - (aic_new2_model$deviance + (aic_new2_model$edf - df_0))/D_0,
  1 - (bic_new_model$deviance + (bic_new_model$edf - df_0))/D_0,
  1 - (p2aic_model$deviance + (p2aic_model$edf - df_0))/D_0,
  1 - (p2bic_model$deviance + (p2bic_model$edf - df_0))/D_0),
  digits = 3)

info2
```
```{r}
data |>
  mutate(
    p = predict(excl_model, type = "probs"),
    yhat = predict(excl_model)) -> excl_pred

glimpse(excl_pred)
confusionMatrix(data = excl_pred$yhat, 
                reference = excl_pred$plasma_123)
```
```{r}
data |>
  mutate(
    p = predict(aic_new2_model, type = "probs"),
    yhat = predict(aic_new2_model)) -> aic_new2_pred

glimpse(aic_new2_pred)
confusionMatrix(data = aic_new2_pred$yhat, 
                reference = aic_new2_pred$plasma_123)
```
```{r}
data |>
  mutate(
    p = predict(bic_new_model, type = "probs"),
    yhat = predict(bic_new_model)) -> bic_new_pred

glimpse(bic_new_pred)
confusionMatrix(data = bic_new_pred$yhat, 
                reference = bic_new_pred$plasma_123)
```

```{r}
data |>
  mutate(
    p = predict(p2aic_model, type = "probs"),
    yhat = predict(p2aic_model)) -> p2aic_pred

glimpse(p2aic_pred)
confusionMatrix(data = p2aic_pred$yhat, 
                reference = p2aic_pred$plasma_123)
```
```{r}
data |>
  mutate(
    p = predict(p2bic_model, type = "probs"),
    yhat = predict(p2bic_model)) -> p2bic_pred

glimpse(p2bic_pred)
confusionMatrix(data = p2bic_pred$yhat, 
                reference = p2bic_pred$plasma_123)
```

